Estimating dark matter halo concentration using the integrated mass profile

### 2016-04-17 paper/paper.tex
fixed typos

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.   
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of the new method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
Convergence tests by downsampling halos with high number of particles in
the N-body simulation show that the new method is robust with respect
to changes in halo particle number.
However the velocity method overpredicts concentration values by a
factor of $25\%$ values for low particle numbers.
We also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{10}\hMsun < M_{h} <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
We find a small deviation at low masses, $M<10^{12}\hMsun$, where
the new method yields lower median concentration values by $25\%$
compared to the velocity methods. 
We find that accounting for the bias
in the velocity method brings the results into agreement.
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

There are two stablished algorithms to estimate the concentration
parameter of a N-body dark matter halo.  The first method takes the
halo particles and bins them into logarithmic radii to estimate the
density in each bin, then it proceeds to fit the density as a function
of the radius to the NFW profile.  A second method uses an analytic
property of the NFW profile that relates the maximum of the ratio of
the circular velocity to the virial velocity, $V_{\rm circ}$/$V_{\rm
  vir}$.  The concentration can be then found as the root of an
algebraic equation dependent on this maximum value.

@@

\begin{equation}
\mathrm{max}(v(<x)) = \sqrt{\frac{c}{x_{\rm max}}\frac{f(x_{\rm
      max})}{f(c)}},
\label{eq:max_v}
\end{equation}
where $x_{\rm max}=2.163$ \citep{Klypin2014} and the function $f(x)$
corresponds to the definition in Eq. (\ref{eq:f_NFW}).

@@

A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).
This method has been most recently used by \cite{Klypin2014} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{error.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.
    \label{fig:error}}
\end{figure}

%\label{sec:data}
%\begin{figure*}
%  \begin{center}
%    \includegraphics[width=0.32\textwidth]{conc_mass_vel.pdf}
%    \includegraphics[width=0.32\textwidth]{conc_mass_dens.pdf}
%    \includegraphics[width=0.32\textwidth]{conc_dens_vel.pdf}
%  \end{center}
%\vspace{-0.5cm}
%  \caption{Comparison between the concentrations measured by the new 
%    method and the maximum velocity (left) and density (middle)
%    methods. The solid line indicates the equal value between the two
%    techniques. The dashed line shows the upper 15 per cent difference
%    between two methods.
%  \label{fig:comparison}}
%\end{figure*}

@@

We quantify the difference between the expected $c_{in}$ and measured
$c_{out}$ values by
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$.  A
large discrepancy between the estimated and input concentrations are
quantified by a large $\avg{|D|}$.

@@

We also find that the bias from different concentrations at fixed
particle number is negligeable with respect to the effects induced by
different particle numbers.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.47\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Uncertainties on different concentration estimates.  The
  large symbols show the standard deviation of the fractional offset
  ($f_{\rm off}$ Eq. \ref{eq:f_off}) from the downsampling
  experiments.  The small dots correspond to the $1-\sigma$
  uncertainty on the concentration for each halo as estimated by the
  MCMC algorithm {\bf delete ``in''} used for the new method.
  Uncertainties from fluctuations on the halo particle number are
  larger than the statistical uncertainties from the fitting
  process. {\bf what is the solid line showing?}
  \label{fig:downsampling_err}}
\end{figure}

\subsection{Measuring the concentration in halos from a cosmological simulation}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data.  All halos are
  used regardless of their relaxation state.  The lines correspond to
  the median concentration values in each mass bin.  The shaded region
  presents 10 and 90 per cent spread.  For clarity we only show the
  spread for the concentrations estimated using the new method. The
  other two methods have a similar spread. The dotted line corresponds
  to fits reported by \citep{Prada2012}.
  \label{fig:concentration}. The left panel shows the raw results
  comming from each algorithm. The right panel introduces a
  correction on the velocity and integrated mass results. 
  The corrections dependens on the number of particles following the
  results of the downsampling experiments presented in Figure
  \ref{fig:downsampling}.} 
\end{figure*}

@@

We perform first a convergence test on the N-body data to find
possible biases in the concentration estimates as a function of
particle number.  We select $8$ halos from the simulation that have at
least $3\times 10^5$ particles.  We then {\bf randomly} downsample the
total number of particles in each halo by factors of $10, 20, 45, 100,
215, 460, 1000$ to measure its concentration with both the integrated
mass and the velocity method. {\bf For each} downsampling factor we
repeat the operation $100$ times.

The average concentration value for the largest number of particles,
$c_{N_{max}}$, provides with a baseline to compare all the other
results.  We use the following statistic
\begin{equation}
f_{\rm off} = c_{N} / c_{N_{max}} - 1,
\end{equation}
to account for the offset between the concentration at a given
downsampled particle number $c_{N}$ and the baseline $c_{N_{max}}$.

Figure \ref{fig:downsampling} summarizes our results.  The plot shows
the average value of $f_{\rm off}$ as a function of the particle
number.  As expected for large enough sample particle numbers,
$N_{s}>5\times 10^3$, the results of the two algorithms coincide and
$\langle f_{\rm off}\rangle\sim 0$.  For lower number of particles,
the results of the two algorithms deviate from the expected
concentration value.  The velocity method overestimates by a factor
$\langle f_{\rm off}\rangle\sim 0.25$ the concentration values for the
lowest sample particle numbers, $N_{s}\sim 100$.  Around the same
sampling scale, the new algorithm shows a more stable behaviour with
an understimate by a factor of $\langle f_{\rm off}\rangle \sim 0.02$.
This shows that in the more complex setup of a cosmological {\bf
  simulation} our algorithm continues to be stable below the $5\%$
level for low particle numbers.

The line on the same panel shows a fit to the data with the following
functional form 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}}.
\label{eq:f_off}
\end{equation} 
for the dataset shown in the panel we have $A=3385\pm 2065$, $B=7.99\pm0.48
$ for the velocity method and $A=-110\pm 253$, $B=7.11\pm1.80$ for the
new method, where the uncertainties are estimated from the covariance 
matrix resulting from the non-linear optimization procedure. The lines
on Figure \ref{fig:downsampling} are plotted using these best values.

Figure \ref{fig:downsampling} {\bf Aca no iría una referencia Fig 3 en
  lugar de Fig 2?} summarizes different uncertainty results.  Large
symbols show the standard deviation on $f_{\rm off}$ from the
downsampling experiment.  Small dots show the uncertainty on each
individual concentration value as estimated by the MCMC algorithm.
The line shows the $\propto N^{-1}$ locus.  This figure demonstrates
that the statistical uncertainties are at least one order of magnitude
{\bf smaller} than the uncertainty associated to downsampling
experiment.

@@

We notice first that the results from the density method have a
systematic $15\%$offset from the velocity methods.  This offset was
already presented by \citep{Prada2012} for some concentrations and
halo masses, although a more organized study in terms of the
mass-concentration relationship has not been presented in the
literature.

We then notice that our {\bf method} follows closely the results of
the velocity method at higher masses, $M_h>10^{12}\hMsun$ or
equivalently for $>5\times10^3$ particles.  For masses below that
threshold our method saturates at a concentration value of $\sim 7$
while the velocity method continues to give higher concentration
values.

We hypothesize that the increase in the results for the velocity
method below $5\times 10^{4}$ particles come from the systematic bias
described in the previous subsection.  To test the general consistency
of this hypothesis {\bf we} correct the concentration values in the
velocity and integrated mass methods by a factor of $1/(1+f_{\rm
  off})$, using the definition in Equation (\ref{eq:f_off}) and the
parameters obtained from the data presented in Figure
(\ref{fig:downsampling}).  The correction brings into perfect
agreement the results between the velocity and the integrated mass
method.  This suggests that the concentration at lower masses might
very well reach a plateu in the halo mass range $10^{10}< M_{h}/\hMsun
< 10^{12}$ .

@@

We also applied the method to halos extracted from a cosmological
N-body simulation.  Using this data we showed that the new method is
robust with respect to a decreasing number of particles in the halos.
On the other hand we showed that the velocity method has a systematic
bias that overesimates the concentration for halos with a low particle
number up to a factor of $25\%$.

{\bf So far no procedure could provide a direct way to measure the
  bias in the estimated halo concentration at low particle numbers and
  a way to use it to make corrections on the estimates. In section 4.2
  it was shown a simple method that can be used on any dataset to
  estimate the errors and correct them when estimating the
  concentration at low particle numbers.}

We also used the N-body simulation to estimate the impact on the mass
concentration relationship.  Although the three methods are in broad
agreement within their uncertainties there are some noticeable
differences.  The first difference is that the density method produces
systematically higher concentrations by a factor of $15\%$ compared to
the velocity mehod.  The second difference is that the velocity and
integrated mass methods start to differ for masses below
$10^{12}\hMsun$.  We found that a correction of the results by the
bias factor found using the same N-body data can bring these two
methods into agreement.

These are encouraging results.  They show that using the integrated
mass profile to find the DM halo concentration is a tool deserving
deeper scrutiny.  Further tests with larger simulated volumes, varying
numerical resolution and different density profiles are the next
natural step to explore the full potential of this new method.

@@

\bibliographystyle{mn2e}
%\bibliography{references}
\begin{thebibliography}{}

@@

\bibitem[\protect\citeauthoryear{{Klypin}, {Yepes}, {Gottlober}, {Prada} \&
  {Hess}}{{Klypin} et~al.}{2014}]{Klypin2014}
{Klypin} A.,  {Yepes} G.,  {Gottlober} S.,  {Prada} F.,    {Hess} S.,  2014,
  ArXiv e-prints

>>>>

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.   
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of the new method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
We design another simple test that can be used on any
data-set to estimate and correct the biases when estimating the
concentration at low particle numbers. 
That test shows that the new method is robust with respect to changes
in halo particle number. 
However, the velocity method over-predicts concentration values by a
factor of $25\%$ values for low particle numbers.
Finally, we also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{10}\hMsun < M_{h} <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
There is a small deviation at low masses, $M<10^{12}\hMsun$, where
the new method yields lower median concentration values by $25\%$
compared to the velocity methods; accounting for the bias in the
velocity method brings the results into agreement.  
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

There are two established algorithms to estimate the concentration
parameter of a N-body dark matter halo.  The first method takes the
halo particles and bins them into logarithmic radii to estimate the
density in each bin, then it proceeds to fit the density as a function
of the radius to the NFW profile.  A second method uses an analytic
property of the NFW profile that relates the maximum of the ratio of
the circular velocity to the virial velocity, $V_{\rm circ}$/$V_{\rm
  vir}$.  The concentration can be then found as the root of an
algebraic equation dependent on this maximum value.

@@

\begin{equation}
\mathrm{max}(v(<x)) = \sqrt{\frac{c}{x_{\rm max}}\frac{f(x_{\rm
      max})}{f(c)}},
\label{eq:max_v}
\end{equation}
where $x_{\rm max}=2.163$ \citep{Klypin2016} and the function $f(x)$
corresponds to the definition in Eq. (\ref{eq:f_NFW}).

@@

A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).
This method has been most recently used by \cite{Klypin2016} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{error.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.
    \label{fig:error}}
\end{figure}

@@

We quantify the difference between the expected $c_{in}$ and measured
$c_{out}$ values by
%
\begin{equation}
D= c_{out}/c_{in}-1,
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$.  A
large discrepancy between the estimated and input concentrations are
quantified by a large $\avg{|D|}$.

@@

We also find that the bias from different concentrations at fixed
particle number is negligible with respect to the effects induced by
different particle numbers.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.47\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Uncertainties on different concentration estimates.  The
  large symbols show the standard deviation of the fractional offset
  ($f_{\rm off}$ as defined in Eq. \ref{eq:f_off}) from the downsampling
  experiments.  The small dots correspond to the $1\sigma$
  uncertainty on the concentration for each halo as estimated by the
  MCMC algorithm used for the new method.
  The solid lines shows the $1/N$ tendency, which also corresponds
  to the particle number dependence of the $\langle|D|\rangle$ statistic
  (i.e. the accuracy to retrieve a known concentration in mock halos)
  of the new method.
  Uncertainties due to particle number fluctuations are always larger
  than the statistical uncertainties from the fitting process. 
  \label{fig:downsampling_err}}
\end{figure}

\subsection{Measuring the concentration in halos from a cosmological simulation}
\label{subsec:cosmo}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data.  All halos are
  used regardless of their relaxation state.  The lines correspond to
  the median concentration values in each mass bin.  The shaded region
  presents 10 and 90 per cent spread.  For clarity we only show the
  spread for the concentrations estimated using the new method. The
  other two methods have a similar spread. The dotted line corresponds
  to fits reported by \citep{Prada2012}.
  \label{fig:concentration}. The left panel shows the raw results
  coming from each algorithm. The right panel introduces a
  correction on the velocity and integrated mass results. 
  The corrections depends on the number of particles following the
  results of the downsampling experiments presented in Figure
  \ref{fig:downsampling}.} 
\end{figure*}

@@

We perform first a convergence test on the N-body data to find
possible biases in the concentration estimates as a function of
particle number.  
We select $8$ halos from the simulation that have at
least $3\times 10^5$ particles.  
We then randomly downsample the total number of particles in each halo
by factors of $10, 20, 45, 100, 215, 460, 1000$ to measure its
concentration with both the integrated mass and the velocity
method. For each downsampling factor we repeat the operation $100$
times.

The average concentration value for the largest number of particles,
$c_{N_{max}}$, provides with a baseline to compare all the other
results.  
We use the following statistic
\begin{equation}
f_{\rm off} = c_{N} / c_{N_{max}} - 1,
\label{eq:f_off}
\end{equation}
to account for the offset between the concentration at a given
downsampled particle number $c_{N}$ and the baseline $c_{N_{max}}$.

Figure \ref{fig:downsampling} summarizes our results.  The plot shows
the average value of $f_{\rm off}$ as a function of the particle
number.  As expected for large enough sample particle numbers,
$N_{s}>5\times 10^3$, the results of the two algorithms coincide and
$\langle f_{\rm off}\rangle\sim 0$.  For lower number of particles,
the results of the two algorithms deviate from the expected
concentration value.  The velocity method overestimates by a factor
$\langle f_{\rm off}\rangle\sim 0.25$ the concentration values for the
lowest sample particle numbers, $N_{s}\sim 100$.  Around the same
sampling scale, the new algorithm shows a more stable behaviour with
an underestimate by a factor of $\langle f_{\rm off}\rangle \sim 0.02$.
This shows that in the more complex setup of a cosmological 
  simulation the new algorithm continues to be stable below the $5\%$
level for low particle numbers.

The line on the same panel shows a fit to the data with the following
functional form 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}}.
\label{eq:f_off_fit}
\end{equation} 
for the dataset shown in the panel we have $A=3385\pm 2065$, $B=7.99\pm0.48
$ for the velocity method and $A=-110\pm 253$, $B=7.11\pm1.80$ for the
new method, where the uncertainties are estimated from the covariance 
matrix resulting from the non-linear optimization procedure. The lines
on Figure \ref{fig:downsampling} are plotted using these best values.

Figure \ref{fig:downsampling_err} summarizes different uncertainty
results.  
Large symbols show the standard deviation on $f_{\rm off}$ from the
downsampling experiment.  
Small dots show the uncertainty on each individual concentration value
as estimated by the MCMC algorithm. 
The line shows the $\propto N^{-1}$ locus.  
This figure demonstrates that the statistical uncertainties are at
least one order of magnitude smaller than the uncertainty associated
to downsampling experiment.

@@

We notice first that the results from the density method have a
systematic $15\%$offset from the velocity methods.  
This offset was already presented by \citep{Prada2012} for low
concentrations ($c<6$) and high ($M_h>10^{12}\hMsun$) halo masses.  
Recently \citep{Klypin2016} summarized results for the
mass-concentration relationship coming from different methods and
datasets to show that similar systematic offsets are present.
However, to the best of our knowledge, a systematic and modern study of
the mass-concentration relationship spanning a wide range in mass,
using a fixed dataset but different fitting methods has not been
presented in the literature.

We then notice that the new method closely follows the results of
the velocity method at higher masses, $M_h>10^{12}\hMsun$ or
equivalently for $>5\times10^3$ particles.  
For masses below that threshold the new method saturates at a
concentration value of $\sim 8$ while the velocity method continues to
give higher concentration values.

We hypothesize that the increase in the results for the velocity
method below $5\times 10^{4}$ particles come from the systematic bias
described in the previous subsection.  To test the general consistency
of this hypothesis we correct the concentration values in the
velocity and integrated mass methods by a factor of $1/(1+f_{\rm
  off})$, using the definition in Equation (\ref{eq:f_off}) and the 
parameters obtained from the data presented in Figure
(\ref{fig:downsampling}).  
The correction brings into perfect agreement the results between the
velocity and the integrated mass method.  
This suggests that the concentration at lower masses might very well
reach a plateau in the halo mass range $10^{10}< M_{h}/\hMsun <
10^{12}$ .

@@

We also applied the method to halos extracted from a cosmological
N-body simulation.  Using this data we showed that the new method is
robust with respect to a decreasing number of particles in the halos.
On the other hand we showed that the velocity method has a systematic
bias that overestimates the concentration for halos with a low particle
number up to a factor of $25\%$.

We provide for the first time in the literature a direct way to
measure the bias in the estimated halo concentration at low particle numbers and
a way to use it to make corrections on the estimates. 
In Section \ref{subsec:cosmo} we presented a simple method that can be used on any
dataset to estimate the errors and correct them when estimating the
concentration at low particle numbers.

We also used the N-body simulation to estimate the impact on the mass
concentration relationship.  
Although the three methods are in broad agreement within their
uncertainties there are some noticeable differences.  
The first difference is that the density method produces
systematically higher concentrations by a factor of $15\%$ compared to
the velocity method.
This systematic offset has been reported in different datasets and
techniques to measure concentrations, without any conclusive
explanation for its origin \citep{Klypin2016}.
The second difference is that the velocity and integrated mass methods
start to differ for masses below $10^{12}\hMsun$.  
We found that a correction of the results by the bias factor found
using the same N-body data can bring these two methods into
agreement.

These are encouraging results.  
They show that using the integrated mass profile to find the Dark
Matter halo concentration is a tool deserving deeper scrutiny.  
Further tests with larger simulated volumes, varying numerical
resolution and different density profiles are the next natural step to
explore the full potential of this new method.

@@

\bibliographystyle{mn2e}
\bibliography{references}

\begin{thebibliography}{}

@@

\bibitem[\protect\citeauthoryear{{Klypin}, {Yepes}, {Gottl{\"o}ber}, {Prada} \&
  {He{\ss}}}{{Klypin} et~al.}{2016}]{Klypin2016}
{Klypin} A.,  {Yepes} G.,  {Gottl{\"o}ber} S.,  {Prada} F.,    {He{\ss}} S.,
  2016, \mnras, 457, 4340
